---
title: "range_limits_brms"
author: "Shahar Chaikin"
date: "2024-05-06"
output: html_document
---

Using brms to answer:
1. How does max. temperature relates with relative polewardness? One would excpect that higher max. temp. will increase the probability of occurrence in the poleward side - hence positive association.

2. How does min. temperature relates with relative polewardness? One would excpect that lower min. temp. will reduce the probability of occurrence in the poleward side - hence negative association.


*add sqrt occurrences as weights
*add biomes and species/ order as random effects

Packages
```{r}
library(tidyverse)
library(brms)
```

Data
```{r}
data=read.csv("Results_all_sps(in).csv") %>% 
  mutate(abs_lat=abs(decimalLatitude),
         sqrt_n_occ=sqrt(n_occ)) %>% #add abs lat
  select(species,
         key,
         abs_lat,
         rangeSize,
         distEdge,
         relPolarwardness,
         elevation,
         biome,
         bodyMass,
         sqrt_n_occ,
         n_occ,
         order,
         Min_T_SHAP,
         Mean_T_SHAP,
         Max_T_SHAP,
         Min_PPT_SHAP,
         Mean_PPT_SHAP,
         Max_PPT_SHAP,
         NOTE) %>% 
    filter(!NOTE%in%"Crossed by Equator",
           n_occ>=20)
data$biome=as.factor(data$biome)
data$species %>% n_distinct()
```

#Polewardness
##max_temp - prior 1
```{r}
max_temp_model=brm(
  data = data,
  formula = Max_T_SHAP~relPolarwardness+(1+relPolarwardness|biome)+(1+relPolarwardness|species),
  chains=3,
  iter = 4000,
  warmup = 2000,
  cores = 3,
  prior = set_prior('normal(0,5)',
                    class="b",
                    coef="relPolarwardness"),
  family=gaussian())#22 min
#save model
saveRDS(max_temp_model, file = "max_temp_model.RDS") 

brms::bayes_R2(max_temp_model)
summary(max_temp_model)
max_temp_model_gg=ggeffects::ggpredict(max_temp_model,terms="relPolarwardness[all]")

#Diagnosis
brms::pp_check(max_temp_model,ndraws=99)# posterior predictive checks
plot(max_temp_model)
```
plot
```{r}
ggplot()+
  geom_point(data=data
             ,aes(x=relPolarwardness,y=Min_T_SHAP),
             alpha=0.3,
             size=1,
             color='grey')+
  geom_ribbon(data=max_temp_model_gg,aes(x=x,ymin=conf.low,ymax=conf.high),
              alpha=0.3,
              fill="orange")+
    geom_line(data=max_temp_model_gg,aes(x=x,y=predicted))+
  labs(x="Relative polewardness",
       y="Max. temp (SHAP)")+
  theme_bw()
```

##max_temp - prior 2
```{r}
max_temp_model_2=brm(
  data = data,
  formula = Max_T_SHAP~relPolarwardness+(1+relPolarwardness|biome)+(1+relPolarwardness|species),
  chains=3,
  iter = 4000,
  warmup = 2000,
  cores = 3,
  prior = set_prior('normal(1,5)',
                    class="b",
                    coef="relPolarwardness"),
  family=gaussian())#22 min
#save model
saveRDS(max_temp_model_2, file = "max_temp_model_2.RDS") 

brms::bayes_R2(max_temp_model_2)
summary(max_temp_model_2)
max_temp_model_gg=ggeffects::ggpredict(max_temp_model_2,terms="relPolarwardness[all]")

#Diagnosis
brms::pp_check(max_temp_model_2,ndraws=99)# posterior predictive checks
plot(max_temp_model_2)
```

##min_temp
```{r}
min_temp_model=brm(
  data = data,
  formula = Min_T_SHAP~relPolarwardness+(1+relPolarwardness|biome)+(1+relPolarwardness|species),
  chains=3,
  iter = 4000,
  warmup = 2000,
  cores = 3,
  prior = set_prior('normal(0,5)',
                    class="b",
                    coef="relPolarwardness"),
  family=gaussian())#22 min
#save model
saveRDS(min_temp_model, file = "min_temp_model.RDS") 

brms::bayes_R2(min_temp_model)
summary(min_temp_model)
min_temp_model_gg=ggeffects::ggpredict(min_temp_model,terms="relPolarwardness[all]")
#Diagnosis
#Diagnosis
brms::pp_check(min_temp_model,ndraws=99)# posterior predictive checks
plot(min_temp_model)
```

Plot
```{r}
ggplot()+
  geom_point(data=data %>% 
               filter(!NOTE%in%"Crossed by Equator",
             n_occ>=20)
             ,aes(x=relPolarwardness,y=Min_T_SHAP),
             alpha=0.3,
             size=1,
             color='grey')+
  geom_ribbon(data=min_temp_model_gg,aes(x=x,ymin=conf.low,ymax=conf.high),
              alpha=0.3,
              fill="blue")+
    geom_line(data=min_temp_model_gg,aes(x=x,y=predicted))+
  labs(x="Relative polewardness",
       y="Min. temp (SHAP)")+
  theme_bw()
```

#Polewardness & elevation
##max_temp - 18 hours
We assume that minimum temperature will limit the poleward expansion of species and that maximum temperature will facilitate it. Hence, prior are built accordingly and are partially informative.
```{r}
#get priors
max_temp_ele_prior=get_prior(Max_T_SHAP~relPolarwardness*elevation+(1|order), data = data)
#reduce complexity by having more simple random effects
#model
max_temp_ele_model=brm(
  data = data,
  formula = Max_T_SHAP~relPolarwardness*elevation,
  chains=3,
  iter = 25000,
  warmup = 2500,
  cores = 3,
  family=gaussian())
get_prior(max_temp_ele_model)
# ,
#   prior = c(set_prior('normal(0.5,10)',
#                     class="b",
#                     coef="relPolarwardness"),#Consider making relPolarwardness positive with large space for variation as we expect positive association
#             set_prior('snormal(-0.5,10)',
#                     class="b",
#                     coef="elevation"),##Consider making elevation negative with large space for variation as we expect negative association
#             set_prior('normal(0,10)',
#                     class="b",
#                     coef="relPolarwardness:elevation")))

#save model
saveRDS(max_temp_ele_model, file = "max_temp_ele_model_5.RDS") 
#max_temp_ele_model=read_rds("max_temp_ele_model_set_prior.RDS")
brms::bayes_R2(max_temp_ele_model)
summary(max_temp_ele_model)
max_temp_ele_model_gg=ggeffects::ggpredict(max_temp_ele_model,terms="relPolarwardness[all]")


#Diagnosis
brms::pp_check(max_temp_ele_model,ndraws=99)# posterior predictive checks
plot(max_temp_ele_model)
```
plot
```{r}
max_temp_ele_model_p=sjPlot::plot_model(max_temp_ele_model,type="pred",terms=c("relPolarwardness","elevation[0,3000,6000]"))+
  geom_vline(xintercept = 0.5,linetype="dashed")+
  labs(x="Relative polewardness",
       y="Max. temp. SHAP",
       color="Elevation (m)")+
   geom_rug(data=data,
           aes(x=relPolarwardness,y=Max_T_SHAP),
           alpha = 0.2,
           linewidth=.5,
           inherit.aes = FALSE)+
  theme_bw()


dens1 <- ggplot(data, aes(x = relPolarwardness)) + 
  geom_density(alpha = 0.4,fill="grey") + 
  theme_void() + 
  theme(legend.position = "none")

dens2 <- ggplot(data, aes(x = Max_T_SHAP)) + 
  geom_density(alpha = 0.4,fill="grey") + 
  theme_void() + 
  theme(legend.position = "none") + 
  coord_flip()
library(ggpubr)
library(patchwork)

max_temp_ele_model_p_dans=dens1 + plot_spacer() + max_temp_ele_model_p + dens2 + 
  plot_layout(ncol = 2, nrow = 2, widths = c(4, 1), heights = c(1, 4))

ggsave(plot = max_temp_ele_model_p_dans,filename = "max_temp_ele_model_p_dans.png",device = "jpeg",units = "cm",width = 20,height = 12)
```

##min_temp
```{r}
min_temp_model=brm(
  data = data,
  formula = Min_T_SHAP~relPolarwardness*elevation,
  chains=3,
  iter = 25000,
  warmup = 2500,
  cores = 3,
  family=gaussian())
#save model
saveRDS(min_temp_model, file = "min_temp_model_ele.RDS") 

brms::bayes_R2(min_temp_model)
summary(min_temp_model)
min_temp_model_gg=ggeffects::ggpredict(min_temp_model,terms="relPolarwardness[all]")
sjPlot::plot_model(min_temp_model,type="int",terms=c("relPolarwardness","elevation[0,6000]"))
#Diagnosis
#Diagnosis
brms::pp_check(min_temp_model,ndraws=99)# posterior predictive checks
plot(min_temp_model)
```

Plot
plot
```{r}
min_temp_ele_model_p=sjPlot::plot_model(min_temp_model,type="pred",terms=c("relPolarwardness","elevation[0,3000,6000]"))+
  geom_vline(xintercept = 0.5,linetype="dashed")+
  labs(x="Relative polewardness",
       y="Min. temp. SHAP",
       color="Elevation (m)")+
   geom_rug(data=data,
           aes(x=relPolarwardness,y=Min_T_SHAP),
           alpha = 0.2,
           linewidth=.5,
           inherit.aes = FALSE)+
  theme_bw()


dens1_min_temp_ele_model_p <- ggplot(data, aes(x = relPolarwardness)) + 
  geom_density(alpha = 0.4,fill="grey") + 
  theme_void() + 
  theme(legend.position = "none")

dens2_min_temp_ele_model_p <- ggplot(data, aes(x = Min_T_SHAP)) + 
  geom_density(alpha = 0.4,fill="grey") + 
  theme_void() + 
  theme(legend.position = "none") + 
  coord_flip()

min_temp_ele_model_p_dans=dens1_min_temp_ele_model_p + plot_spacer() + min_temp_ele_model_p + dens2_min_temp_ele_model_p + 
  plot_layout(ncol = 2, nrow = 2, widths = c(4, 1), heights = c(1, 4))

ggsave(plot = min_temp_ele_model_p_dans,filename = "min_temp_ele_model_p_dans.png",device = "jpeg",units = "cm",width = 20,height = 12)
```



Test the contribution of Shap values at different:
1. relative polewardness
2. elevation
3. dist. from edge
